---
title: "ADNI-Analysis9-2"
output: html_document
date: "2024-12-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("dplyr")
library("ggplot2")
library("tidyr")
library("purrr")
library("dtwclust")
library("xgboost")
library("caTools")
library("caret")
library("pROC")
```

The goal of this report is to plot the latent space result based on the xGradient boost on previous meeting

```{r}
data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/Meeting-9/ADNIMERGE-dtw.csv")
```

```{r}
train_xgb_model <- function(data) {
  # Ensure correct levels for `cluster`
  data$cluster <- factor(data$cluster, levels = c("Stable", "Mild Progression", "Moderate Progression", "Severe Progression"))
  
  # Split data
  set.seed(123)
  sample_split <- sample.split(Y = data$cluster, SplitRatio = 0.7)
  train_set <- subset(data, sample_split == TRUE)
  test_set <- subset(data, sample_split == FALSE)
  
  # Prepare labels and features
  y_train <- as.integer(train_set$cluster) - 1
  y_test <- as.integer(test_set$cluster) - 1
  X_train <- train_set %>% dplyr::select(-cluster) %>% as.data.frame() %>% mutate_all(as.numeric)
  X_test <- test_set %>% dplyr::select(-cluster) %>% as.data.frame() %>% mutate_all(as.numeric)
  
  # Convert to xgb.DMatrix
  xgb_train <- xgb.DMatrix(data = as.matrix(X_train), label = y_train)
  xgb_test <- xgb.DMatrix(data = as.matrix(X_test), label = y_test)
  
  # Set parameters
  num_classes <- length(unique(y_train))  # Ensure correct num_class
  xgb_params <- list(
    booster = "gbtree",
    eta = 0.01,
    max_depth = 8,
    gamma = 4,
    subsample = 0.75,
    colsample_bytree = 1,
    objective = "multi:softprob",
    eval_metric = "mlogloss",
    num_class = num_classes
  )
  
  # Train the model
  xgb_model <- xgb.train(
    params = xgb_params,
    data = xgb_train,
    nrounds = 5000,
    verbose = 1
  )
  
  # Predict and calculate AUC
  pred_probs <- predict(xgb_model, newdata = xgb_test)
  pred_probs <- matrix(pred_probs, ncol = num_classes, byrow = TRUE)
  auc_values <- sapply(0:(num_classes - 1), function(class) {
    binary_labels <- ifelse(y_test == class, 1, 0)
    roc_obj <- roc(binary_labels, pred_probs[, class + 1])
    auc(roc_obj)
  })
  
  names(auc_values) <- paste("AUC_Class", 0:(num_classes - 1), sep = "_")
  
  # Return the model and AUC
  return(list(model = xgb_model, auc = auc_values))
}

```

First one we see the CG XGBoost of the cognitive test, and then we will take the 2 most significant variables and plot it in the next one.

```{r}
cg_data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/Meeting-9/cg_latent.csv")
cg_data$cluster <- factor(cg_data$cluster, 
                           levels = c(0, 1, 2, 3), 
                           labels = c("Mild Progression", "Moderate Progression", "Severe Progression","Stable"))
cg_xgb <- train_xgb_model(cg_data)
cg_importance_matrix <- xgb.importance(model = cg_xgb$model)
print(cg_importance_matrix)
xgb.plot.importance(cg_importance_matrix)
```

As we can see the most significance are latent 4 and 1, so we will try to plot that.

```{r}
ggplot(cg_data, aes(x = latent_4, y = latent_1, color = as.factor(cluster))) +
  geom_point() +
  labs(title = "Latent Space: Latent_4 vs Latent_1 (Clustered)",
       x = "Latent_4",
       y = "Latent_1",
       color = "Cluster") +
  theme_minimal()
```

To draw meaningful conclusions, we can perform linear regression to explore the relationships between the latent space, cluster results, and key variables such as age, gender, and APOE4. From the analysis, it is evident that age is not well-represented in the latent space. 

```{r}
cg_data <- cg_data %>% dplyr:::select(-cluster)
cg_sets <- cbind(data, cg_data)
cg_sets <- cg_sets[!is.na(cg_sets$AGE) & !is.infinite(cg_sets$AGE), ]
cg_sets$cluster_set <- as.numeric(factor(cg_sets$cluster))
model <- lm(AGE ~ latent_1 + latent_4 + cluster_set, data = cg_sets)
summary(model)
```

Gender shows a slightly stronger association but remains limited.

```{r}
cg_sets$gender <- as.numeric(factor(cg_sets$PTGENDER))
model <- lm(gender ~ latent_1 + latent_4 + cluster_set, data = cg_sets)
summary(model)
```

In contrast, APOE4 demonstrates significant relevance, highlighting its critical role in understanding Alzheimer's patients and its strong representation in the latent space.

```{r}
model <- lm(APOE4 ~ latent_1 + latent_4 + cluster_set, data = cg_sets)
summary(model)
```

Next one we are moving on to the CSF test

```{r}
csf_data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/Meeting-9/csf_latent.csv")
csf_data$cluster <- factor(csf_data$cluster, 
                           levels = c(0, 1, 2, 3), 
                           labels = c("Mild Progression", "Moderate Progression", "Severe Progression","Stable"))
csf_xgb <- train_xgb_model(csf_data)
csf_importance_matrix <- xgb.importance(model = csf_xgb$model)
print(csf_importance_matrix)
xgb.plot.importance(csf_importance_matrix)
```

```{r}
all_features <- colnames(csf_data)[colnames(csf_data) != "cluster"]

# Ensure all features are in the importance matrix
missing_features <- setdiff(all_features, csf_importance_matrix$Feature)

# Add missing features with zero importance
if (length(missing_features) > 0) {
  zero_importance <- data.frame(
    Feature = missing_features,
    Gain = 0,
    Cover = 0,
    Frequency = 0,
    Importance = 0
  )
  csf_importance_matrix <- rbind(csf_importance_matrix, zero_importance)
}

xgb.plot.importance(csf_importance_matrix)
```

As we can see the most significance are latent 4 and 1, so we will try to plot that.

```{r}
ggplot(csf_data, aes(x = latent_4, y = latent_1, color = as.factor(cluster))) +
  geom_point() +
  labs(title = "Latent Space: Latent_4 vs Latent_1 (Clustered)",
       x = "Latent_4",
       y = "Latent_1",
       color = "Cluster") +
  theme_minimal()
```

We can see similar result in csf set with the cognitive set. So we can see that age and gender is not really captured well, but APOE4 is actually pretty significant.

```{r}
csf_data <- csf_data %>% dplyr:::select(-cluster)
csf_sets <- cbind(data, csf_data)
csf_sets$cluster_set <- as.numeric(factor(csf_sets$cluster))
model <- lm(AGE ~ latent_1 + latent_4 + cluster_set, data = csf_sets)
summary(model)
```

```{r}
csf_sets$gender <- as.numeric(factor(csf_sets$PTGENDER))
model <- lm(gender ~ latent_1 + latent_4 + cluster_set, data = csf_sets)
summary(model)
```

```{r}
model <- lm(APOE4 ~ latent_1 + latent_4 + cluster_set, data = csf_sets)
summary(model)
```

Next we will move to the PET set

```{r}
pet_data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/Meeting-9/pet_latent.csv")
pet_data$cluster <- factor(pet_data$cluster, 
                           levels = c(0, 1, 2, 3), 
                           labels = c("Mild Progression", "Moderate Progression", "Severe Progression","Stable"))
pet_xgb <- train_xgb_model(pet_data)
pet_importance_matrix <- xgb.importance(model = pet_xgb$model)
print(pet_importance_matrix)
xgb.plot.importance(pet_importance_matrix)
```

```{r}
all_features <- colnames(pet_data)[colnames(pet_data) != "cluster"]

# Ensure all features are in the importance matrix
missing_features <- setdiff(all_features, pet_importance_matrix$Feature)

# Add missing features with zero importance
if (length(missing_features) > 0) {
  zero_importance <- data.frame(
    Feature = missing_features,
    Gain = 0,
    Cover = 0,
    Frequency = 0,
    Importance = 0
  )
  pet_importance_matrix <- rbind(pet_importance_matrix, zero_importance)
}

xgb.plot.importance(pet_importance_matrix)
```

As we can see the most significance are latent 1 and 2, so we will try to plot that.

```{r}
ggplot(pet_data, aes(x = latent_1, y = latent_2, color = as.factor(cluster))) +
  geom_point() +
  labs(title = "Latent Space: Latent_1 vs Latent_2 (Clustered)",
       x = "Latent_1",
       y = "Latent_2",
       color = "Cluster") +
  theme_minimal()
```

```{r}
pet_data <- pet_data %>% dplyr:::select(-cluster)
pet_sets <- cbind(data, pet_data)
pet_sets$cluster_set <- as.numeric(factor(pet_sets$cluster))
model <- lm(AGE ~ latent_1 + latent_2 + cluster_set, data = pet_sets)
summary(model)
```

Comparing these results to previous models, it becomes evident that the representation of AGE in the latent space remains consistently weak, with a small R² and limited predictor significance, much like earlier findings. Gender shows some improvement, with all predictors (latent_1, latent_2, and cluster_set) statistically significant in this model, but the explanatory power (R² = 0.01098) is still minimal, consistent with earlier conclusions of weak alignment.

```{r}
pet_sets$gender <- as.numeric(factor(csf_sets$PTGENDER))
model <- lm(gender ~ latent_1 + latent_2 + cluster_set, data = pet_sets)
summary(model)
```

The APOE4 model demonstrates a stronger relationship with the latent space and cluster set compared to age and gender. With an R² of 0.05627 and a highly significant contribution from cluster_set, the latent space appears to effectively capture biological information relevant to APOE4. In contrast, age and gender remain poorly represented, reinforcing the idea that the latent space primarily encodes features tied to Alzheimer's-related biomarkers like APOE4.

```{r}
model <- lm(APOE4 ~ latent_1 + latent_2 + cluster_set, data = pet_sets)
summary(model)
```

Then we are moving on to use the risk set

```{r}
risk_data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/Meeting-9/risk_latent.csv")
risk_data$cluster <- factor(risk_data$cluster, 
                           levels = c(0, 1, 2, 3), 
                           labels = c("Mild Progression", "Moderate Progression", "Severe Progression","Stable"))
risk_xgb <- train_xgb_model(risk_data)
risk_importance_matrix <- xgb.importance(model = risk_xgb$model)
print(risk_importance_matrix)
xgb.plot.importance(risk_importance_matrix)
```

```{r}
all_features <- colnames(risk_data)[colnames(risk_data) != "cluster"]

# Ensure all features are in the importance matrix
missing_features <- setdiff(all_features, risk_importance_matrix$Feature)

# Add missing features with zero importance
if (length(missing_features) > 0) {
  zero_importance <- data.frame(
    Feature = missing_features,
    Gain = 0,
    Cover = 0,
    Frequency = 0,
    Importance = 0
  )
  risk_importance_matrix <- rbind(risk_importance_matrix, zero_importance)
}

xgb.plot.importance(risk_importance_matrix)
```

As we can see the most significance are latent 1 and 2, so we will try to plot that.

```{r}
ggplot(risk_data, aes(x = latent_4, y = latent_1, color = as.factor(cluster))) +
  geom_point() +
  labs(title = "Latent Space: Latent_4 vs Latent_1 (Clustered)",
       x = "Latent_4",
       y = "Latent_1",
       color = "Cluster") +
  theme_minimal()
```

We also applied this analysis to the risk set, and the results showed that all variables—gender, age, and APOE4—were statistically significant. This outcome is expected, as these variables were incorporated into the creation of the latent space, introducing potential collinearity between the predictors and the latent dimensions.

```{r}
risk_data <- risk_data %>% dplyr:::select(-cluster)
risk_sets <- cbind(data, risk_data)
risk_sets$cluster_set <- as.numeric(factor(risk_sets$cluster))
model <- lm(AGE ~ latent_1 + latent_4 + cluster_set, data = risk_sets)
summary(model)
```

```{r}
risk_sets$gender <- as.numeric(factor(risk_sets$PTGENDER))
model <- lm(gender ~ latent_1 + latent_4 + cluster_set, data = risk_sets)
summary(model)
```

```{r}
model <- lm(APOE4 ~ latent_1 + latent_4 + cluster_set, data = risk_sets)
summary(model)
```

The next step is MRI set

```{r}
mri_data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/Meeting-9/mri_latent.csv")
mri_data$cluster <- factor(mri_data$cluster, 
                           levels = c(0, 1, 2, 3), 
                           labels = c("Mild Progression", "Moderate Progression", "Severe Progression","Stable"))
mri_xgb <- train_xgb_model(mri_data)
mri_importance_matrix <- xgb.importance(model = mri_xgb$model)
print(mri_importance_matrix)
xgb.plot.importance(mri_importance_matrix)
```

```{r}
all_features <- colnames(mri_data)[colnames(mri_data) != "cluster"]

# Ensure all features are in the importance matrix
missing_features <- setdiff(all_features, mri_importance_matrix$Feature)

# Add missing features with zero importance
if (length(missing_features) > 0) {
  zero_importance <- data.frame(
    Feature = missing_features,
    Gain = 0,
    Cover = 0,
    Frequency = 0,
    Importance = 0
  )
  mri_importance_matrix <- rbind(mri_importance_matrix, zero_importance)
}

xgb.plot.importance(mri_importance_matrix)
```

Because of the MRI set only has latent 3 value, we will skip to plot this part. But we will do the linear regression to find the relation between latent space and age, gender, and APOE4 variables.

The MRI set shows a moderate relationship between the latent space and APOE4, with the clustering structure (cluster_set) being a key predictor. Gender is better captured than AGE, as seen in its stronger predictor significance and slightly higher R². However, both AGE and APOE4 remain less robustly represented compared to gender, suggesting that the latent space better encodes demographic features like gender than complex biological markers like APOE4.

```{r}
mri_data <- mri_data %>% dplyr:::select(-cluster)
mri_sets <- cbind(data, mri_data)
mri_sets$cluster_set <- as.numeric(factor(mri_sets$cluster))
model <- lm(AGE ~ latent_3  + cluster_set, data = mri_sets)
summary(model)
```

```{r}
mri_sets$gender <- as.numeric(factor(mri_sets$PTGENDER))
model <- lm(gender ~ latent_3 + cluster_set, data = mri_sets)
summary(model)
```

```{r}
model <- lm(APOE4 ~ latent_3 + cluster_set, data = mri_sets)
summary(model)
```

The last step we will try to do the same step with mri and risk set

```{r}
mri_risk_data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/Meeting-9/mri_risk_latent.csv")
mri_risk_data$cluster <- factor(mri_risk_data$cluster, 
                           levels = c(0, 1, 2, 3), 
                           labels = c("Mild Progression", "Moderate Progression", "Severe Progression","Stable"))
mri_risk_xgb <- train_xgb_model(mri_risk_data)
mri_risk_importance_matrix <- xgb.importance(model = mri_risk_xgb$model)
print(mri_risk_importance_matrix)
xgb.plot.importance(mri_risk_importance_matrix)
```

We will use latent 2 and 3 on this case

```{r}
ggplot(mri_risk_data, aes(x = latent_2, y = latent_3, color = as.factor(cluster))) +
  geom_point() +
  labs(title = "Latent Space: Latent_2 vs Latent_3 (Clustered)",
       x = "Latent_2",
       y = "Latent_3",
       color = "Cluster") +
  theme_minimal()
```

On this case also makes sense because we use risk set, so there is a potentital of collinearity.

```{r}
mri_risk_data <- mri_risk_data %>% dplyr:::select(-cluster)
mri_risk_sets <- cbind(data, mri_risk_data)
mri_risk_sets$cluster_set <- as.numeric(factor(mri_risk_sets$cluster))
model <- lm(AGE ~ latent_2 + latent_3  + cluster_set, data = mri_risk_sets)
summary(model)
```

```{r}
mri_risk_sets$gender <- as.numeric(factor(mri_risk_sets$PTGENDER))
model <- lm(gender ~ latent_2 + latent_3 + cluster_set, data = mri_risk_sets)
summary(model)
```

```{r}
model <- lm(APOE4 ~ latent_2 + latent_3 + cluster_set, data = mri_risk_sets)
summary(model)
```