---
title: "ADNI-Analysis7"
output: html_document
date: "2024-11-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("dplyr")
library("ggplot2")
library("tidyr")
library("purrr")
library("dtwclust")
library("xgboost")
library("caTools")
library("caret")
library("pROC")
```

## R Markdown

```{r}
normalize_minmax <- function(x) {
  return ((x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE)))
}

sort_viscode <- function(viscode, months, cdrsb) {
  visit_numeric <- as.numeric(gsub("m", "", gsub("bl", "0", viscode)))
  sorted_indices <- order(visit_numeric)
  list(
    viscode_sorted = viscode[sorted_indices],
    cdrsb_sorted = cdrsb[sorted_indices],
    month_sorted = months[sorted_indices]
  )
}
```

```{r}
train_xgb_model <- function(data) {
  # Split data
  data$cluster <- factor(data$cluster, levels = c("Stable", "Mild Progression", "Moderate Progression", "Severe Progression"))
  set.seed(123)
  sample_split <- sample.split(Y = data$cluster, SplitRatio = 0.7)
  train_set <- subset(data, sample_split == TRUE)
  test_set <- subset(data, sample_split == FALSE)
  
  # Prepare training and testing labels and features
  y_train <- as.integer(train_set$cluster) - 1
  y_test <- as.integer(test_set$cluster) - 1
  X_train <- train_set %>% dplyr::select(-cluster)
  X_test <- test_set %>% dplyr::select(-cluster)
  
  print(levels(train_set$cluster))
  
  # Convert to xgb.DMatrix
  xgb_train <- xgb.DMatrix(data = as.matrix(X_train), label = y_train)
  xgb_test <- xgb.DMatrix(data = as.matrix(X_test), label = y_test)
  
  # Set XGBoost parameters
  xgb_params <- list(
    booster = "gbtree",
    eta = 0.01,
    max_depth = 8,
    gamma = 4,
    subsample = 0.75,
    colsample_bytree = 1,
    objective = "multi:softprob",
    eval_metric = "mlogloss",
    num_class = length(levels(data$cluster))
  )
  
  # Train the model
  xgb_model <- xgb.train(
    params = xgb_params,
    data = xgb_train,
    nrounds = 5000,
    verbose = 1
  )
  
  pred_probs <- predict(xgb_model, newdata = xgb_test)
  num_classes <- xgb_params$num_class
  pred_probs <- matrix(pred_probs, ncol = num_classes, byrow = TRUE)
  auc_values <- sapply(0:(num_classes - 1), function(class) {
    # Binary labels: 1 for current class, 0 for others
    binary_labels <- ifelse(y_test == class, 1, 0)
    roc_obj <- roc(binary_labels, pred_probs[, class + 1])
    auc(roc_obj)
  })
  
  # Print and return the AUC values for each class
  names(auc_values) <- paste("AUC_Class", 0:(num_classes - 1), sep = "_")
  
  # Return the model and accuracy
  return(list(model = xgb_model, auc = auc_values))
}
```

```{r}
# Get fluctuating data
data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/ADNIMERGE-Simple.csv")

data <- data %>%
  filter(!is.na(WholeBrain))

ptid_counts <- table(data$PTID)

ptid_more_than_once <- names(ptid_counts[ptid_counts > 1])
filtered_data <- data %>%
  filter(data$PTID %in% ptid_more_than_once)

grouped_data <- filtered_data %>%
  drop_na(CDRSB) %>%                
  group_by(PTID) %>%
  summarize(
    code = list(VISCODE),
    months = list(Month),
    cdrsb_scores = list(CDRSB)
  ) %>%
  mutate(
    sorted = pmap(list(code, months, cdrsb_scores), sort_viscode), 
    code = map(sorted, "viscode_sorted"),             
    cdrsb_scores = map(sorted, "cdrsb_sorted"),
    months = map(sorted, "month_sorted")
  ) %>%
  select(-sorted) %>%  
  filter(
    !sapply(cdrsb_scores, function(x) all(x == 0))
  )

summary_table <- grouped_data %>%
  mutate(
    cdrsb_mean = map_dbl(cdrsb_scores, mean, na.rm = TRUE),         
    cdrsb_first = map_dbl(cdrsb_scores, ~ .x[1]),                   
    cdrsb_last = map_dbl(cdrsb_scores, ~ .x[length(.x)]),           
    cdrsb_sd = map_dbl(cdrsb_scores, sd, na.rm = TRUE),
    cdrsb_scores = cdrsb_scores,
    months = months,
    visits = map_int(cdrsb_scores, length)
  ) %>%
  select(PTID, cdrsb_mean, cdrsb_first, cdrsb_last, cdrsb_sd, cdrsb_scores, months, visits)

threshold_diff <- 1
threshold_sd <- 1
fluctuating_data <- summary_table %>%
  filter(abs(cdrsb_first - cdrsb_last) < threshold_diff & cdrsb_sd > threshold_sd)

plot_data <- fluctuating_data %>%
  unnest(c(cdrsb_scores, months))  # Unnest the list columns

ggplot(plot_data, aes(x = months, y = cdrsb_scores, group = PTID, color = PTID)) +
  geom_line() +
  labs(title = "CDRSB Progression Over Time by Patient",
       x = "Months", y = "CDRSB Score") +
  theme_minimal() +
  theme(legend.position = "none")  # Hide legend for individual patients
```

```{r}
# Load data
data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/ADNIMERGE-Simple.csv")

norm_data <- data %>% mutate(across(where(is.numeric), normalize_minmax))

ptid_counts <- table(data$PTID)

ptid_more_than_once <- names(ptid_counts[ptid_counts > 1])
filtered_data <- norm_data %>%
  filter(norm_data$PTID %in% ptid_more_than_once)

grouped_data <- filtered_data %>%
  drop_na(CDRSB) %>%                
  group_by(PTID) %>%
  summarize(
    code = list(VISCODE),
    months = list(Month),
    cdrsb_scores = list(CDRSB)
  ) %>%
  mutate(
    sorted = pmap(list(code, months, cdrsb_scores), sort_viscode), 
    code = map(sorted, "viscode_sorted"),             
    cdrsb_scores = map(sorted, "cdrsb_sorted"),
    months = map(sorted, "month_sorted")
  ) %>%
  dplyr::select(-sorted) %>%  
  filter(
    !sapply(cdrsb_scores, function(x) all(x == 0))
  )
```

```{r}
# DTW
time_series_data <- grouped_data$cdrsb_scores

dtw_kmeans_result <- tsclust(time_series_data, 
                             type = "partitional",  # Partitional clustering (k-means)
                             k = 4,                 # Number of clusters (adjust k as needed)
                             distance = "dtw",      # Use DTW as distance metric
                             centroid = "pam",      # PAM-based centroid (good for non-Euclidean)
                             seed = 123)            # Seed for reproducibility

cluster_assignments <- dtw_kmeans_result@cluster
dtw_data <- grouped_data
dtw_data$cluster <- as.factor(cluster_assignments)
dtw_data <- dtw_data %>%
  unnest(cols = c(months, cdrsb_scores))

dtw_data <- dtw_data %>%
  mutate(cluster = factor(cluster, 
                          levels = c(1, 2, 3, 4), 
                          labels = c("Mild Progression", 
                                     "Stable", 
                                     "Severe Progression", 
                                     "Moderate Progression")))

```

We can see below the DTW of the normalized data that we did before.

```{r}
# make sure to take only PTID and cluster
filtered_dtw <- dtw_data %>%
  dplyr::select(PTID, cluster) %>%
  group_by(PTID) %>%
  dplyr::slice(1)

merged_data <- left_join(filtered_dtw, data, by = "PTID")
write.csv(merged_data, "/Users/nathaniel.putera/Projects/UCPH/MRI-Project/ADNIMERGE-Simple-dtw.csv", row.names = FALSE)
```

```{r}
ggplot(dtw_data, aes(x = months, y = cdrsb_scores, group = PTID, color = factor(cluster))) +
  geom_line(alpha = 0.6) +
  labs(title = "CDRSB Progression by Cluster",
       x = "Months of Visit",
       y = "CDRSB Score",
       color = "Cluster") +
  theme_bw()
```

```{r}
fluctuating_clusters <- fluctuating_data %>%
  inner_join(dtw_data %>% select(PTID, cluster), by = "PTID")

fluctuating_clusters %>% select(PTID, cdrsb_first, cdrsb_last, cdrsb_mean, cdrsb_sd, cluster) %>% distinct()
```

With this DTW cluster result, we will try to use it as a label for the classification later. So after merging the cluster result with the original data, we will only take the first occurence only then do XGBoost on the data to find which variables is actually significance to the result. We will make couple of sets, and will compare the result using AUC to see which one is better.

```{r}
# This to filter to show only 1 data
first_occurrence <- merged_data %>%
  group_by(PTID) %>% 
  dplyr::slice(1) %>%
  ungroup()
```

Below are the first xgboost we do using cognitive test set.

```{r}
# Set using Cognitive test
cg_set <- first_occurrence %>%
  dplyr::select(c(FAQ, RAVLT_learning, RAVLT_forgetting, RAVLT_immediate, MMSE, ADAS11, ADAS13, ADASQ4, LDELTOTAL, DIGITSCOR, TRABSCOR, MOCA, cluster))

cg_xgb <- train_xgb_model(cg_set)
cg_importance_matrix <- xgb.importance(model = cg_xgb$model)
print(cg_importance_matrix)
xgb.plot.importance(cg_importance_matrix)
```

Below are the next xgboost we do using CSF (Cerebrospinal fluid) set.

```{r}
# Set using CSF (Cerebrospinal fluid)
csf_set <- first_occurrence %>%
  mutate(
    ABETA = as.numeric(gsub(">1700", "2500", ABETA)),
    TAU = as.numeric(gsub(">1700", "1700", TAU)),
    PTAU = as.numeric(gsub(">1700", "1700", PTAU))
  ) %>%
  select(c(ABETA, TAU, PTAU, cluster))

csf_xgb <- train_xgb_model(csf_set)
csf_importance_matrix <- xgb.importance(model = csf_xgb$model)
print(csf_importance_matrix)
xgb.plot.importance(csf_importance_matrix)
```

Below are the next xgboost we do using PET (Positron Emission Tomography) set.

```{r}
# Set using PET (Positron Emission Tomography)
pet_set <- first_occurrence %>%
  select(c(FDG, PIB, AV45, FBB, cluster))

pet_xgb <- train_xgb_model(pet_set)
pet_importance_matrix <- xgb.importance(model = pet_xgb$model)
print(pet_importance_matrix)
xgb.plot.importance(pet_importance_matrix)
```

Below are the next xgboost we do using risk factor set.

```{r}
# Set using risk factor
risk_set <- first_occurrence %>%
  select(c(AGE, PTGENDER, APOE4, PTEDUCAT, cluster)) %>%
  mutate(PTGENDER = ifelse(PTGENDER == "Male", 1, 0))

risk_xgb <- train_xgb_model(risk_set)
risk_importance_matrix <- xgb.importance(model = risk_xgb$model)
print(risk_importance_matrix)
xgb.plot.importance(risk_importance_matrix)
```

Below are the next xgboost we do using MRI Biomarker set.

```{r}
# Set using MRI
mri_set <- first_occurrence %>%
  mutate(
    Ventricles = (Ventricles / ICV),
    Hippocampus = (Hippocampus / ICV),
    WholeBrain = (WholeBrain / ICV),
    Entorhinal = (Entorhinal / ICV),
    Fusiform = (Fusiform / ICV),
    MidTemp = (MidTemp / ICV))%>%
  select(c(Ventricles, Hippocampus, WholeBrain, Entorhinal, Fusiform, MidTemp, cluster))

mri_xgb <- train_xgb_model(mri_set)
mri_importance_matrix <- xgb.importance(model = mri_xgb$model)
print(mri_importance_matrix)
xgb.plot.importance(mri_importance_matrix)
```

Next we want to compare how each section perform by comparing the AUC.

```{r}
# This is not valid anymore as we change it to use auc values instead accuracy, figure this out later
auc_data <- data.frame(
  Model = rep(c("cg_xgb", "csf_xgb", "mri_xgb", "risk_xgb", "pet_xgb"), each = 4),
  Class = rep(c("Class_0", "Class_1", "Class_2", "Class_3"), times = 5),
  AUC = c(cg_xgb$auc, csf_xgb$auc, mri_xgb$auc, risk_xgb$auc, pet_xgb$auc)
)

class_mapping <- c("Class_0" = "Stable", "Class_1" = "Mild Progression", "Class_2" = "Moderate Progression", "Class_3" = "Severe Progression")
auc_data$Class <- class_mapping[auc_data$Class]

print(auc_data$AUC)

ggplot(auc_data, aes(x = Model, y = AUC, fill = Class)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "AUC Comparison Across Models and Classes", x = "Model", y = "AUC") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")
```

Lastly doing comparison using MRI and Risk as it is the easiest data to collect

```{r}
mri_risk_set <- first_occurrence %>%
  mutate(
    Ventricles = (Ventricles / ICV),
    Hippocampus = (Hippocampus / ICV),
    WholeBrain = (WholeBrain / ICV),
    Entorhinal = (Entorhinal / ICV),
    Fusiform = (Fusiform / ICV),
    MidTemp = (MidTemp / ICV))%>%
  select(c(Ventricles, Hippocampus, WholeBrain, Entorhinal, Fusiform, MidTemp, AGE, PTGENDER, APOE4, PTEDUCAT, cluster)) %>%
  mutate(PTGENDER = ifelse(PTGENDER == "Male", 1, 0))

mri_risk_xgb <- train_xgb_model(mri_risk_set)
mri_risk_importance_matrix <- xgb.importance(model = mri_risk_xgb$model)
print(mri_risk_importance_matrix)
xgb.plot.importance(mri_risk_importance_matrix)
```


Then we can see the AUC values of this set as below

```{r}
mri_risk_xgb$auc
```