---
title: "ADNI-Analysis2"
output: html_document
date: "2024-09-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("dplyr")
library("ggplot2")
library("stringr")
library("lmtest")
library("stargazer")
library("purrr")
library("tidyr")
```

## Part 3 ADNI Analysis

```{r cars}
data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/ADNIMERGE-Simple.csv")

ptid_counts <- table(data$PTID)

data <- data %>%
  filter(!is.na(WholeBrain)) %>%
  mutate(
    Ventricles_percent = (Ventricles / ICV),
    Hippocampus_percent = (Hippocampus / ICV),
    WholeBrain_percent = (WholeBrain / ICV),
    Entorhinal_percent = (Entorhinal / ICV),
    Fusiform_percent = (Fusiform / ICV),
    MidTemp_percent = (MidTemp / ICV)
  )

ptid_more_than_once <- names(ptid_counts[ptid_counts > 1])
filtered_data <- data %>%
  filter(data$PTID %in% ptid_more_than_once)
```

```{r}
ggplot(data, aes(x = Month, y = CDRSB, group = PTID, color = PTID)) +
  geom_line(alpha = 0.6) +  # Connect points with lines for each patient
  geom_point(alpha = 0.6) + # Add points for each visit
  labs(title = "Progression of Cognitive Decline Over Time (CDRSB)",
       x = "Month of Visit",
       y = "CDRSB Score") +
  theme_minimal() +
  theme(legend.position = "none") 
```

So here for the first part, I want to try to show how is it look like progress of the CDRSB value over visit for all patients, of course on the plot above we cannot see anything, so we can try to do clustering and maybe color it by cluster.

```{r}
sort_viscode <- function(viscode, months, cdrsb, brainVolumes, hippocampus, ventricles) {
  visit_numeric <- as.numeric(gsub("m", "", gsub("bl", "0", viscode)))
  sorted_indices <- order(visit_numeric)
  list(
    viscode_sorted = viscode[sorted_indices],
    cdrsb_sorted = cdrsb[sorted_indices],
    month_sorted = months[sorted_indices],
    brain_volumes = brainVolumes[sorted_indices],
    hippocampus_volumes = hippocampus[sorted_indices],
    ventricles_volumes = ventricles[sorted_indices]
  )
}

grouped_data <- filtered_data %>%
  drop_na(CDRSB) %>%                
  group_by(PTID) %>%
  summarize(
    code = list(VISCODE),
    months = list(Month),
    cdrsb_scores = list(CDRSB),
    brain_volumes = list(WholeBrain_percent),
    hippocampus_volumes = list(Hippocampus_percent),
    ventricles_volumes = list(Ventricles_percent)
  ) %>%
  mutate(
    sorted = pmap(list(code, months, cdrsb_scores, brain_volumes, hippocampus_volumes, ventricles_volumes), sort_viscode), 
    code = map(sorted, "viscode_sorted"),             
    cdrsb_scores = map(sorted, "cdrsb_sorted"),
    months = map(sorted, "month_sorted"),
    brain_volumes = map(sorted, "brain_volumes"),
    hippocampus_volumes = map(sorted, "hippocampus_volumes"),
    ventricles_volumes = map(sorted, "ventricles_volumes"),
  ) %>%
  select(-sorted) %>%  
  filter(
    !sapply(cdrsb_scores, function(x) all(x == 0))
  )

summary_table <- grouped_data %>%
  mutate(
    cdrsb_mean = map_dbl(cdrsb_scores, mean, na.rm = TRUE),         
    cdrsb_first = map_dbl(cdrsb_scores, ~ .x[1]),                   
    cdrsb_last = map_dbl(cdrsb_scores, ~ .x[length(.x)]),           
    cdrsb_sd = map_dbl(cdrsb_scores, sd, na.rm = TRUE)              
  ) %>%
  select(PTID, cdrsb_mean, cdrsb_first, cdrsb_last, cdrsb_sd)

cleaned_summary_table <- summary_table %>%
  drop_na()

set.seed(123)
kmeans_result <- kmeans(cleaned_summary_table %>% select(-PTID), centers = 4)

cleaned_summary_table <- cleaned_summary_table %>%
  mutate(cluster = kmeans_result$cluster)

grouped_data_with_clusters <- grouped_data %>%
  left_join(cleaned_summary_table %>% select(PTID, cluster), by = "PTID")

long_data <- grouped_data_with_clusters %>%
  unnest(c(months, cdrsb_scores, brain_volumes, hippocampus_volumes, ventricles_volumes))

long_data <- long_data %>%
  mutate(cluster = factor(cluster, 
                          levels = c(1, 2, 3, 4), 
                          labels = c("Mild Progression", 
                                     "Moderate Progression", 
                                     "Severe Progression", 
                                     "Stable")))

# Plot the CDRSB progression for each cluster
ggplot(long_data, aes(x = months, y = cdrsb_scores, group = PTID, color = factor(cluster))) +
  geom_line(alpha = 0.6) +
  labs(title = "CDRSB Progression by Cluster",
       x = "Months of Visit",
       y = "CDRSB Score",
       color = "Cluster") +
  theme_bw()
```

Here, I used 4 clusters with the aim of capturing different progression patterns: stable, moderate, fast progression, and a cluster for people whose CDRSB values fluctuated significantly. Although the result differs slightly from what I expected, we can still observe that each cluster is grouped distinctly.

The data I used to feed into the K-means algorithm includes the CDRSB values from the first and last visits for each patient, along with the mean and standard deviation of the CDRSB scores. This approach helps capture not only the overall progression but also the variability between visits. I couldn't use all CDRSB values directly since each patient has a different number of visits, leading to inconsistent data lengths across patients. Despite these limitations, the clustering still reveals meaningful groupings.

Looking at the plot, we can see how patients with varying progression rates are distributed across the clusters, providing insight into different patterns of cognitive decline.

```{r}
# Summary of each clusters
summary_by_patient <- long_data %>%
  group_by(PTID, cluster) %>%
  summarize(
    first_cdrsb = first(cdrsb_scores),   # First visit CDRSB
    last_cdrsb = last(cdrsb_scores),     # Last visit CDRSB
    mean_cdrsb = mean(cdrsb_scores, na.rm = TRUE)  # Mean CDRSB across all visits
  )

# Now calculate the mean and standard deviation for each cluster
summary_by_cluster <- summary_by_patient %>%
  group_by(cluster) %>%
  summarize(
    mean_start_cdrsb = mean(first_cdrsb, na.rm = TRUE),  # Mean first visit CDRSB
    mean_last_cdrsb = mean(last_cdrsb, na.rm = TRUE),    # Mean last visit CDRSB
    mean_of_means_cdrsb = mean(mean_cdrsb, na.rm = TRUE), # Mean of all patient means within the cluster
  )

summary_by_cluster

```

We can see the progress from each cluster above. With the stable really show almost no progression, mild progression a bit from start mean 1.86 to 3.84, then moderate from 3.87 to 6.55 and lastly severe has the biggest jump from 4.31 to 11.89.

```{r}
long_data_grouped <- long_data %>%
  group_by(cluster, PTID) %>%
  arrange(months)

# Brain volumes each clusters
ggplot(long_data_grouped, aes(x = months, y = brain_volumes, group = PTID, color = PTID)) +
  geom_line() +
  labs(title = "Brain Volume Change Over Time by Cluster",
       x = "Months", y = "Brain Volume") +
  theme_minimal() +
  theme(legend.position = "none") +  # Hide legend for individual patients
  facet_wrap(~ cluster, scales = "free_y")  # Create one plot per cluster with free y-axis scaling

```

With each cluster plotted for whole brain volume changes, we can observe how brain size is impacted over time. The mild progression cluster shows relatively stable trends with slower brain volume loss, while the moderate progression cluster displays a gradual but noticeable decline, consistent with brain atrophy over time. The severe progression cluster, however, exhibits a sharper and more rapid decrease in brain volume, indicative of more aggressive disease progression. In contrast, the stable cluster shows minimal changes in brain volume, potentially representing individuals with slower or non-progressing cognitive impairment. These patterns offer valuable insight into how different levels of disease progression affect brain volume, potentially aiding in understanding the relationship between brain atrophy and cognitive decline in Alzheimer's disease.

```{r}
hippocampus_grouped <- long_data_grouped %>%
  filter(!is.na(hippocampus_volumes))

ggplot(hippocampus_grouped, aes(x = months, y = ventricles_volumes, group = PTID, color = PTID)) +
  geom_line() +
  labs(title = "Hippocampus Volume Change Over Time by Cluster",
       x = "Months", y = "Hippocampus Volume") +
  theme_minimal() +
  theme(legend.position = "none") +  # Hide legend for individual patients
  facet_wrap(~ cluster, scales = "free_y")  # Create one plot per cluster with free y-axis scaling
```

On the hippocampus volume plot, we observe a notable trend where the volume appears to increase, particularly in the more severe progression clusters. In the mild progression and stable clusters, the hippocampus volume generally shows a smaller degree of change, remaining relatively steady over time. However, as we move to the moderate and severe progression clusters, there is a more pronounced increase in hippocampus volume. This could indicate that as cognitive decline accelerates, certain brain regions exhibit compensatory volume changes or potentially reflect measurement variability. These trends highlight how brain volume changes differ across the severity of disease progression, particularly in a key region like the hippocampus, which is often associated with memory and cognitive functions.

```{r}
ventricles_grouped <- long_data_grouped %>%
  filter(!is.na(ventricles_volumes))

ggplot(long_data_grouped, aes(x = months, y = ventricles_volumes, group = PTID, color = PTID)) +
  geom_line() +
  labs(title = "Ventricles Volume Change Over Time by Cluster",
       x = "Months", y = "Ventricles Volume") +
  theme_minimal() +
  theme(legend.position = "none") +  # Hide legend for individual patients
  facet_wrap(~ cluster, scales = "free_y")  # Create one plot per cluster with free y-axis scaling
```

The same thing also happens in the ventricles, with more severe progression have more increase.

```{r}
ptid_cluster_dict <- long_data %>%
  distinct(PTID, cluster) 

merged_data <- data %>%
  inner_join(ptid_cluster_dict, by = "PTID") %>%
  distinct(PTID, APOE4, DX, cluster)

# APOE4
apoe4_summary <- merged_data %>%
  group_by(cluster, APOE4) %>%
  summarize(count = n())  # Count how many patients have each APOE4 status in each cluster

apoe4_summary_table <- apoe4_summary %>%
  group_by(cluster) %>%
  summarize(
    apoe4_0 = sum(count[APOE4 == 0], na.rm = TRUE),  # Summing up APOE4=0 count
    apoe4_1 = sum(count[APOE4 == 1], na.rm = TRUE),  # Summing up APOE4=1 count
    apoe4_2 = sum(count[APOE4 == 2], na.rm = TRUE),  # Summing up APOE4=2 count
    apoe4_na = sum(count[is.na(APOE4)], na.rm = TRUE)  # Summing up APOE4 missing values (NA)
  )

print(apoe4_summary_table)
```

The summary reveals interesting trends in the distribution of APOE4 status across clusters. In the Stable group, there is a noticeably higher proportion of patients with APOE4 = 0 compared to the other clusters, indicating that a larger share of this group does not carry the APOE4 allele. In contrast, the Mild Progression group has a more balanced distribution across the APOE4 statuses, with similar proportions of APOE4 = 0 and APOE4 = 1 patients, but a smaller proportion of APOE4 = 2. The Moderate Progression cluster tends to have more patients with APOE4 = 1, while the Severe Progression cluster has a higher proportion of APOE4 = 1 carriers compared to those with APOE4 = 0. Overall, it appears that the Stable group has more non-carriers, while clusters showing greater cognitive decline tend to have a larger proportion of patients with APOE4 alleles, particularly APOE4 = 1.

```{r}
dx_summary <- merged_data %>%
  group_by(cluster, DX) %>%
  summarize(count = n())  # Count how many patients have each APOE4 status in each cluster

dx_summary_table <- dx_summary %>%
  group_by(cluster) %>%
  summarize(
    ad = sum(count[DX == "AD"], na.rm = TRUE), 
    cn = sum(count[DX == "CN"], na.rm = TRUE),  
    mci = sum(count[DX == "MCI"], na.rm = TRUE), 
    na = sum(count[is.na(DX)], na.rm = TRUE)  
  )

print(dx_summary_table)
```

We can also see the distribution of each cluster based on its diagnosis, where the stable group has the most cognitive normal people and less people diagnosed alzheimer. While on more severe progress like moderate and severe group, we can see the distribution of the patients show more on the "AD" group compare to other 2 groups.