---
title: "ADNI-Analysis10"
output: html_document
date: "2025-01-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("dplyr")
library("ggplot2")
library("tidyr")
library("purrr")
library("dtwclust")
library("xgboost")
library("caTools")
library("caret")
library("pROC")
library("plotly")
```

## R Markdown
We will redo the work we did in the Analysis 9-2. Instead of using 2d plot, we will use 3d plot, and also include all latent in the linear regression.

```{r}
data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/Meeting-9/ADNIMERGE-dtw.csv")
```

```{r}
train_xgb_model <- function(data) {
  # Ensure correct levels for `cluster`
  data$cluster <- factor(data$cluster, levels = c("Stable", "Mild Progression", "Moderate Progression", "Severe Progression"))
  
  # Split data
  set.seed(123)
  sample_split <- sample.split(Y = data$cluster, SplitRatio = 0.7)
  train_set <- subset(data, sample_split == TRUE)
  test_set <- subset(data, sample_split == FALSE)
  
  # Prepare labels and features
  y_train <- as.integer(train_set$cluster) - 1
  y_test <- as.integer(test_set$cluster) - 1
  X_train <- train_set %>% dplyr::select(-cluster) %>% as.data.frame() %>% mutate_all(as.numeric)
  X_test <- test_set %>% dplyr::select(-cluster) %>% as.data.frame() %>% mutate_all(as.numeric)
  
  # Convert to xgb.DMatrix
  xgb_train <- xgb.DMatrix(data = as.matrix(X_train), label = y_train)
  xgb_test <- xgb.DMatrix(data = as.matrix(X_test), label = y_test)
  
  # Set parameters
  num_classes <- length(unique(y_train))  # Ensure correct num_class
  xgb_params <- list(
    booster = "gbtree",
    eta = 0.01,
    max_depth = 8,
    gamma = 4,
    subsample = 0.75,
    colsample_bytree = 1,
    objective = "multi:softprob",
    eval_metric = "mlogloss",
    num_class = num_classes
  )
  
  # Train the model
  xgb_model <- xgb.train(
    params = xgb_params,
    data = xgb_train,
    nrounds = 5000,
    verbose = 1
  )
  
  # Predict and calculate AUC
  pred_probs <- predict(xgb_model, newdata = xgb_test)
  pred_probs <- matrix(pred_probs, ncol = num_classes, byrow = TRUE)
  auc_values <- sapply(0:(num_classes - 1), function(class) {
    binary_labels <- ifelse(y_test == class, 1, 0)
    roc_obj <- roc(binary_labels, pred_probs[, class + 1])
    auc(roc_obj)
  })
  
  names(auc_values) <- paste("AUC_Class", 0:(num_classes - 1), sep = "_")
  
  # Return the model and AUC
  return(list(model = xgb_model, auc = auc_values))
}

```

Starting with Cognitive set

```{r}
cg_data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/Meeting-10/cg_latent.csv")
cg_data$cluster <- factor(cg_data$cluster, 
                           levels = c(0, 1, 2, 3), 
                           labels = c("Mild Progression", "Moderate Progression", "Severe Progression","Stable"))
cg_xgb <- train_xgb_model(cg_data)
cg_importance_matrix <- xgb.importance(model = cg_xgb$model)
print(cg_importance_matrix)
xgb.plot.importance(cg_importance_matrix)
```

```{r}
plot_ly(cg_data, 
        x = ~latent_4, 
        y = ~latent_1, 
        z = ~latent_2, 
        color = ~as.factor(cluster), 
        type = "scatter3d", 
        mode = "markers")

```

Cognitive set linear regression with Age

```{r}
cg_data <- cg_data %>% dplyr:::select(-cluster)
cg_sets <- cbind(data, cg_data)
cg_sets <- cg_sets[!is.na(cg_sets$AGE) & !is.infinite(cg_sets$AGE), ]
cg_sets$cluster_set <- as.numeric(factor(cg_sets$cluster))
model <- lm(AGE ~ latent_1+ latent_2 + latent_3 + latent_4 + cluster_set, data = cg_sets)
summary(model)
```

Cognitive set linear regression with Gender

```{r}
cg_sets$gender <- as.numeric(factor(cg_sets$PTGENDER))
model <- lm(gender ~ latent_1+ latent_2 + latent_3 + latent_4 + cluster_set, data = cg_sets)
summary(model)
```

Cognitive set linear regression with APOE4

```{r}
model <- lm(APOE4 ~ latent_1+ latent_2 + latent_3 + latent_4 + cluster_set, data = cg_sets)
summary(model)
```

Moving on to the csf set

```{r}
csf_data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/Meeting-10/csf_latent.csv")
csf_data$cluster <- factor(csf_data$cluster, 
                           levels = c(0, 1, 2, 3), 
                           labels = c("Mild Progression", "Moderate Progression", "Severe Progression","Stable"))
csf_xgb <- train_xgb_model(csf_data)
csf_importance_matrix <- xgb.importance(model = csf_xgb$model)
print(csf_importance_matrix)
xgb.plot.importance(csf_importance_matrix)
```

As there is still 2 latent spaces here, so we won't redo the csf part

```{r}
ggplot(csf_data, aes(x = latent_4, y = latent_1, color = as.factor(cluster))) +
  geom_point() +
  labs(title = "Latent Space: Latent_4 vs Latent_1 (Clustered)",
       x = "Latent_4",
       y = "Latent_1",
       color = "Cluster") +
  theme_minimal()
```

Plotting 3D the PET set

```{r}
pet_data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/Meeting-10/pet_latent.csv")
pet_data$cluster <- factor(pet_data$cluster, 
                           levels = c(0, 1, 2, 3), 
                           labels = c("Mild Progression", "Moderate Progression", "Severe Progression","Stable"))
pet_xgb <- train_xgb_model(pet_data)
pet_importance_matrix <- xgb.importance(model = pet_xgb$model)
print(pet_importance_matrix)
xgb.plot.importance(pet_importance_matrix)
```

```{r}
plot_ly(pet_data, 
        x = ~latent_1, 
        y = ~latent_2, 
        z = ~latent_4, 
        color = ~as.factor(cluster), 
        type = "scatter3d", 
        mode = "markers")

```

Next to the PET regression

```{r}
pet_data <- pet_data %>% dplyr:::select(-cluster)
pet_sets <- cbind(data, pet_data)
pet_sets$cluster_set <- as.numeric(factor(pet_sets$cluster))
model <- lm(AGE ~ latent_1 + latent_2 + latent_4 + cluster_set, data = pet_sets)
summary(model)
```

```{r}
pet_sets$gender <- as.numeric(factor(pet_sets$PTGENDER))
model <- lm(gender ~ latent_1 + latent_2 + latent_4 + cluster_set, data = pet_sets)
summary(model)
```

```{r}
model <- lm(APOE4 ~ latent_1 + latent_2 + latent_4 + cluster_set, data = pet_sets)
summary(model)
```

Moving on to risk set

```{r}
risk_data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/Meeting-10/risk_latent.csv")
risk_data$cluster <- factor(risk_data$cluster, 
                           levels = c(0, 1, 2, 3), 
                           labels = c("Mild Progression", "Moderate Progression", "Severe Progression","Stable"))
risk_xgb <- train_xgb_model(risk_data)
risk_importance_matrix <- xgb.importance(model = risk_xgb$model)
print(risk_importance_matrix)
xgb.plot.importance(risk_importance_matrix)
```

```{r}
plot_ly(risk_data, 
        x = ~latent_1, 
        y = ~latent_3, 
        z = ~latent_4, 
        color = ~as.factor(cluster), 
        type = "scatter3d", 
        mode = "markers")

```

```{r}
mri_risk_data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/Meeting-10/mri_risk_latent.csv")
mri_risk_data$cluster <- factor(mri_risk_data$cluster, 
                           levels = c(0, 1, 2, 3), 
                           labels = c("Mild Progression", "Moderate Progression", "Severe Progression","Stable"))
mri_risk_xgb <- train_xgb_model(mri_risk_data)
mri_risk_importance_matrix <- xgb.importance(model = mri_risk_xgb$model)
print(mri_risk_importance_matrix)
xgb.plot.importance(mri_risk_importance_matrix)
```

```{r}
plot_ly(mri_risk_data, 
        x = ~latent_1, 
        y = ~latent_2, 
        z = ~latent_3, 
        color = ~as.factor(cluster), 
        type = "scatter3d", 
        mode = "markers")

```
