---
title: "ADNI-Analysis8-3"
output: html_document
date: "2024-12-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("dplyr")
library("ggplot2")
library("tidyr")
library("purrr")
library("dtwclust")
library("xgboost")
library("caTools")
library("caret")
library("pROC")
```

## R Markdown

```{r}
data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/ADNIMERGE-dtw.csv")
```

```{r}
train_xgb_model <- function(data) {
  # Ensure correct levels for `cluster`
  data$cluster <- factor(data$cluster, levels = c("Stable", "Mild Progression", "Moderate Progression", "Severe Progression"))
  
  # Split data
  set.seed(123)
  sample_split <- sample.split(Y = data$cluster, SplitRatio = 0.7)
  train_set <- subset(data, sample_split == TRUE)
  test_set <- subset(data, sample_split == FALSE)
  
  # Prepare labels and features
  y_train <- as.integer(train_set$cluster) - 1
  y_test <- as.integer(test_set$cluster) - 1
  X_train <- train_set %>% dplyr::select(-cluster) %>% as.data.frame() %>% mutate_all(as.numeric)
  X_test <- test_set %>% dplyr::select(-cluster) %>% as.data.frame() %>% mutate_all(as.numeric)
  
  # Convert to xgb.DMatrix
  xgb_train <- xgb.DMatrix(data = as.matrix(X_train), label = y_train)
  xgb_test <- xgb.DMatrix(data = as.matrix(X_test), label = y_test)
  
  # Set parameters
  num_classes <- length(unique(y_train))  # Ensure correct num_class
  xgb_params <- list(
    booster = "gbtree",
    eta = 0.01,
    max_depth = 8,
    gamma = 4,
    subsample = 0.75,
    colsample_bytree = 1,
    objective = "multi:softprob",
    eval_metric = "mlogloss",
    num_class = num_classes
  )
  
  # Train the model
  xgb_model <- xgb.train(
    params = xgb_params,
    data = xgb_train,
    nrounds = 5000,
    verbose = 1
  )
  
  # Predict and calculate AUC
  pred_probs <- predict(xgb_model, newdata = xgb_test)
  pred_probs <- matrix(pred_probs, ncol = num_classes, byrow = TRUE)
  auc_values <- sapply(0:(num_classes - 1), function(class) {
    binary_labels <- ifelse(y_test == class, 1, 0)
    roc_obj <- roc(binary_labels, pred_probs[, class + 1])
    auc(roc_obj)
  })
  
  names(auc_values) <- paste("AUC_Class", 0:(num_classes - 1), sep = "_")
  
  # Return the model and AUC
  return(list(model = xgb_model, auc = auc_values))
}

```

Below we can see CG Latent significance

```{r}
# xgb with cg latent and cluster
cg_data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/Meeting-8/cg_latent.csv")
cg_data$cluster <- factor(cg_data$cluster, 
                           levels = c(0, 1, 2, 3), 
                           labels = c("Mild Progression", "Moderate Progression", "Severe Progression","Stable"))
cg_xgb <- train_xgb_model(cg_data)
cg_importance_matrix <- xgb.importance(model = cg_xgb$model)
print(cg_importance_matrix)
xgb.plot.importance(cg_importance_matrix)
```

Below we can see CSF Latent significance

```{r}
# xgb with csf latent and cluster
csf_data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/Meeting-8/csf_latent.csv")
csf_data$cluster <- factor(csf_data$cluster, 
                           levels = c(0, 1, 2, 3), 
                           labels = c("Mild Progression", "Moderate Progression", "Severe Progression","Stable"))
csf_xgb <- train_xgb_model(csf_data)
csf_importance_matrix <- xgb.importance(model = csf_xgb$model)
print(csf_importance_matrix)
xgb.plot.importance(csf_importance_matrix)
```

Below we can see PET Latent significance

```{r}
# xgb with pet latent and cluster
pet_data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/Meeting-8/pet_latent.csv")
pet_data$cluster <- factor(pet_data$cluster, 
                           levels = c(0, 1, 2, 3), 
                           labels = c("Mild Progression", "Moderate Progression", "Severe Progression","Stable"))
pet_xgb <- train_xgb_model(pet_data)
pet_importance_matrix <- xgb.importance(model = pet_xgb$model)
print(pet_importance_matrix)
print(pet_xgb$auc)
xgb.plot.importance(pet_importance_matrix)
```

Below we can see Risk Latent significance

```{r}
# xgb with risk latent and cluster
risk_data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/Meeting-8/risk_latent.csv")
risk_data$cluster <- factor(risk_data$cluster, 
                           levels = c(0, 1, 2, 3), 
                           labels = c("Mild Progression", "Moderate Progression", "Severe Progression","Stable"))
risk_xgb <- train_xgb_model(risk_data)
risk_importance_matrix <- xgb.importance(model = risk_xgb$model)
print(risk_importance_matrix)
xgb.plot.importance(risk_importance_matrix)
```

Below we can see MRI Latent significance

```{r}
# xgb with mri latent and cluster
mri_data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/Meeting-8/mri_latent.csv")
mri_data$cluster <- factor(mri_data$cluster, 
                           levels = c(0, 1, 2, 3), 
                           labels = c("Mild Progression", "Moderate Progression", "Severe Progression","Stable"))
mri_xgb <- train_xgb_model(mri_data)
mri_importance_matrix <- xgb.importance(model = mri_xgb$model)
print(mri_importance_matrix)
xgb.plot.importance(mri_importance_matrix)
```

Below we can see MRI-Risk Latent significance

```{r}
# xgb with mri-risk latent and cluster
mri_risk_data <- read.csv("/Users/nathaniel.putera/Projects/UCPH/MRI-Project/Meeting-8/mri_risk_latent.csv")
mri_risk_data$cluster <- factor(mri_risk_data$cluster, 
                           levels = c(0, 1, 2, 3), 
                           labels = c("Mild Progression", "Moderate Progression", "Severe Progression","Stable"))
mri_risk_xgb <- train_xgb_model(mri_risk_data)
mri_risk_importance_matrix <- xgb.importance(model = mri_risk_xgb$model)
print(mri_risk_importance_matrix)
xgb.plot.importance(mri_risk_importance_matrix)
```

```{r}
# AUC
auc_data <- data.frame(
  Model = rep(c("cg_xgb", "csf_xgb", "pet_xgb", "mri_xgb", "risk_xgb"), each = 4),
  Class = rep(c("Class_0", "Class_1", "Class_2", "Class_3"), times = 5),
  AUC = c(cg_xgb$auc, csf_xgb$auc, pet_xgb$auc, mri_xgb$auc, risk_xgb$auc)
)

class_mapping <- c("Class_0" = "Stable", "Class_1" = "Mild Progression", "Class_2" = "Moderate Progression", "Class_3" = "Severe Progression")
auc_data$Class <- class_mapping[auc_data$Class]

ggplot(auc_data, aes(x = Model, y = AUC, fill = Class)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "AUC Comparison Across Models and Classes", x = "Model", y = "AUC") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")
```

Below are the Cognitive AUC
```{r}
cg_xgb$auc
```

Below are the CSF AUC
```{r}
csf_xgb$auc
```

Below are the PET AUC
```{r}
pet_xgb$auc
```

Below are the Risk AUC
```{r}
risk_xgb$auc
```

Below are the MRI AUC
```{r}
mri_xgb$auc
```

Below are the MRI - Risk AUC
```{r}
mri_risk_xgb$auc
```